---
title: "Getting started with mHapDMR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mHapDMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

**mHapDMR** detects differentially methylated regions using the **M-score**
derived from mHap-format sequencing data.

M-score for region $j$, sample $d$ is defined as:

$$M_{jd} = \frac{S_{jd}}{T_{jd}}, \quad
  S_{jd} = \sum_t N_{tjd} Z_{tjd}, \quad
  T_{jd} = \sum_t N_{tjd}$$

where $N_{tjd}$ is the number of CpG sites on read $t$ and
$Z_{tjd} \in \{0,1\}$ indicates whether the read carries at least one
methylated site.

Differential analysis is performed by DSS (Beta-Binomial model) after
replacing the standard inputs $(m_{jd}, Y_{jd})$ with $(\kappa_{jd}, Y'_{jd})$:

$$\kappa_{jd} = \frac{T_{jd}^2}{\sum_t N_{tjd}^2}, \qquad
  Y'_{jd} = S_{jd} \cdot \frac{T_{jd}}{\sum_t N_{tjd}^2}$$

so that $Y'_{jd}/\kappa_{jd} = M_{jd}$ exactly.

## Installation

```{r install}
BiocManager::install(c("DSS", "GenomicRanges", "IRanges",
                       "S4Vectors", "Rsamtools", "BiocParallel"))
devtools::install_local("mHapDMR")
```

## Data requirements

Both the mHap files and the CpG position file must be bgzipped and
tabix-indexed:

```bash
bgzip sample.mhap && tabix -p bed sample.mhap.gz
bgzip hg19_CpG.txt && tabix -s1 -b2 -e2 hg19_CpG.txt.gz
```

Verify indexes from R:

```{r check}
library(mHapDMR)
check_tabix_index("sample.mhap.gz")
check_tabix_index("hg19_CpG.gz")
```

## Quick start — full pipeline

```{r quickstart}
library(mHapDMR)
library(GenomicRanges)

# Analysis regions (rGR)
rGR <- GRanges(
  seqnames = c("chr1", "chr1", "chr2"),
  ranges   = IRanges(start = c(1000L, 5000L, 2000L),
                     end   = c(1200L, 5300L, 2400L))
)

# Named mHap files — names become sample IDs
mhap_files <- c(
  Control1 = "data/ctrl1.mhap.gz",
  Control2 = "data/ctrl2.mhap.gz",
  Tumor1   = "data/tumor1.mhap.gz",
  Tumor2   = "data/tumor2.mhap.gz"
)

# Experimental design (row order matches names(mhap_files))
design <- data.frame(
  group     = c("Control", "Control", "Tumor", "Tumor"),
  row.names = names(mhap_files)
)

# Run the pipeline
result <- mhap_dmr(
  mhap_files = mhap_files,
  cpg_file   = "ref/hg19_CpG.gz",
  rGR        = rGR,
  design     = design,
  formula    = ~ group,
  coef       = "groupTumor"
)

# Differential M-score results
head(result$dml_result)

# Significant regions (FDR < 0.05)
subset(result$dml_result, fdrs < 0.05)
```

## Step-by-step usage

The pipeline can also be run step by step, giving access to intermediate
objects.

### Step 1 — read mHap records for one region (mHapGR)

```{r step1}
# Single-interval GRanges (gr)
gr <- GRanges("chr1", IRanges(1000L, 1200L))

# mHapGR: one row per unique haplotype
mHapGR <- read_mhap_gr("ctrl1.mhap.gz", "hg19_CpG.gz", gr)
mHapGR          # ranges = read coordinates, mcols = count/ncpg/hap_str/is_methylated
mHapGR$count    # haplotype counts
mHapGR$ncpg     # CpG sites per read (in the region)
```

### Step 2 — compute M-score stats for all regions (rGR)

```{r step2}
rGR_ctrl1 <- mscore_region_stats("ctrl1.mhap.gz", "hg19_CpG.gz", rGR)

# M-score and DSS parameters are stored as metadata columns
rGR_ctrl1$mscore    # M-score per region
rGR_ctrl1$kappa     # effective sample size (κ)
rGR_ctrl1$Y_prime   # effective M-score numerator (Y')
```

By default only these summary statistics are kept. If you also need the raw
per-region haplotypes, pass `keep_mhapgr = TRUE` — but note this retains one
`GRanges` per region, which is memory-intensive for large region sets across
many samples, so leave it off for genome-wide analyses.

```{r step2-keep, eval = FALSE}
rGR_ctrl1 <- mscore_region_stats("ctrl1.mhap.gz", "hg19_CpG.gz", rGR,
                                 keep_mhapgr = TRUE)
mHapGR_list <- attr(rGR_ctrl1, "mHapGR_list")
mHapGR_list[[1]]    # mHapGR for the first region
```

### Step 3 — build BSseq and run the test

```{r step3}
rGR_list <- list(
  Control1 = mscore_region_stats("ctrl1.mhap.gz", "hg19_CpG.gz", rGR),
  Control2 = mscore_region_stats("ctrl2.mhap.gz", "hg19_CpG.gz", rGR),
  Tumor1   = mscore_region_stats("tumor1.mhap.gz", "hg19_CpG.gz", rGR),
  Tumor2   = mscore_region_stats("tumor2.mhap.gz", "hg19_CpG.gz", rGR)
)

bsseq      <- build_bsseq_mscore(rGR_list, min_reads = 5L)
dml_result <- run_dml_test(bsseq, design, formula = ~ group, coef = "groupTumor")
head(dml_result)
```

## Parallel execution

```{r parallel}
library(BiocParallel)

result <- mhap_dmr(
  mhap_files = mhap_files,
  cpg_file   = "ref/hg19_CpG.gz",
  rGR        = rGR,
  design     = design,
  formula    = ~ group,
  coef       = "groupTumor",
  BPPARAM    = MulticoreParam(workers = 4L)
)
```
